Lossy compression of DNA sequencing quality data

### 2015-04-13 manuscript/q-compression.tex
Some small revisions to text.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
\textsc{lap} score can be viewed as a log likelihood score, where a
value closer to 0 is better. We use a reference-based evaluation
script provided by \textsc{gage} that counts single nucleotide
polymorphisms (\textsc{snp}s), relocations, translations, and
inversions. The reference-based metrics are normalized by the length
of the assembly to aid in comparison. For the genome assembly we used
software that makes use quality values in the assembly process:
\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set outperforms all others in all metrics (Table \ref{fig:assembly_ranks}).
%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set using the uncompressed reads outperforms all compression methods in terms of \textsc{lap}, NG50, relocations, translations, and inversions.
Among the compression methods, the
256-, 64-, 128-profile encoding had the highest ranks, followed by QualComp with rates 10 bits/read, 30 bits/read, 100 bits/read, then 7th-degree polynomial regression, followed by the QualComp with rate 6 bits/read, the 2-bin encoding, and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.

@@

In general, the greater the polynomial order, the better overall assembly; however, the 5th-degree polynomial regression performs slightly worse than the 3rd-degree polynomial.
Their respective ranks in terms of N50 and relocations are fairly distant, which lowers the 5th-degree polynomial's overall ranking barely below the 3rd-degree's.
The 1st- and 0-degree polynomial regression methods perform poor in all metrics except assembled bases.
One explanation is that the high error portions of sequences are being marked as high quality, so \textsc{allpaths-lg} is unable to trim or error correct the sequences.  Assembled sequences that overlap maybe unable to align to the errors at the end of the reads, artificially inflating the assembled genome size.

>>>>

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
\textsc{lap} score can be viewed as a log likelihood score, where a
value closer to 0 is better. We use a script provided by \textsc{gage}
reference-based evaluation to count single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to facilitate comparison. For the genome assembly we used software
that makes use quality values in the assembly process:
\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
outperforms all others in all metrics (Table
\ref{fig:assembly_ranks}).
%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
%using the uncompressed reads outperforms all compression methods in
%terms of \textsc{lap}, NG50, relocations, translations, and
%inversions.
Among the compression methods, the 256-, 64-, 128-profile encoding had
the highest ranks, followed by QualComp with rates 10 bits/read, 30
bits/read, 100 bits/read, then 7th-degree polynomial regression,
followed by the QualComp with rate 6 bits/read, the 2-bin encoding,
and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.

@@

In general, the greater the polynomial order, the better overall
assembly; however, the 5th-degree polynomial regression performs
slightly worse than the 3rd-degree polynomial. The respective ranks in
terms of N50 and relocations are fairly distant, which lowers the
overall ranking of the 5th-degree polynomial slightly below that for
3rd-degree polynomial model. The 1st- and 0-degree polynomial
regression methods perform poor in all metrics except assembled bases.
One explanation is that the high error portions of sequences are being
marked as high quality, so \textsc{allpaths-lg} is unable to trim or
error correct the sequences. Assembled sequences that overlap maybe
unable to align to the errors at the end of the reads, artificially
inflating the assembled genome size.
